---
title: "Hands-on Exercise 1B"
subtitle: "Choropleth mapping with tmap"
author: "Zhenhua Liu"
date: "2026-09-11"
body-classes: exercise-page
toc: true
toc-location: right
code-fold: false
code-tools: true
execute:
  echo: true
  warning: false
  message: false
---

## Overview

This exercise builds a choropleth workflow for Singapore's 2024 resident
population. I prepare demographic indicators at subzone level, join them to the
Master Plan 2019 boundaries, and compare mapping choices rather than accepting
the first default map.

::: {.callout-note title="Data note"}
The current official boundary download is GeoJSON and already exposes its
attributes as columns. The original lesson's `rvest` step was needed to unpack
an HTML description stored inside an older KML file. I checked the new file and
did not parse HTML that is no longer present. This keeps the workflow faithful
to the task while avoiding an artificial cleaning step.
:::

## 1. Setup and import

```{r}
#| label: setup
#| include: false
library(sf)
library(tidyverse)
library(tmap)

data_dir <- "hands-on-exercise-1b/data/raw"
```

The download script saves the official boundary and unpacks the SingStat ZIP.
I explicitly exclude the separate notes CSV when selecting the data table.

```{r}
#| label: locate-population-file
population_files <- list.files(
  file.path(data_dir, "singstat-population-2024"),
  pattern = "[.]csv$",
  recursive = TRUE,
  full.names = TRUE
)

population_file <- population_files[
  !startsWith(basename(population_files), "Notes_")
][1]

basename(population_file)
```

```{r}
#| label: import-data
mpsz2019 <- st_read(
  file.path(data_dir, "mp19-subzone.geojson"),
  quiet = TRUE
) |>
  st_make_valid() |>
  st_transform(3414)

population_raw <- read_csv(population_file, show_col_types = FALSE)

tibble(
  dataset = c("MP2019 subzones", "Population records"),
  rows = c(nrow(mpsz2019), nrow(population_raw)),
  columns = c(ncol(mpsz2019), ncol(population_raw))
)
```

```{r}
#| label: inspect-inputs
glimpse(population_raw)
head(st_drop_geometry(mpsz2019)[, 1:7], 3)
```

Inspecting the files first caught an important detail: population is split by
age group, sex and dwelling type. A simple sum of the raw rows is valid only
after grouping by geography and age; otherwise it is easy to lose track of
what each number represents.

## 2. Preparing demographic variables

I use the lesson's three life-stage groups: 0–19 as young residents, 20–64 as
economically active, and 65 or older as aged residents. The dependency ratio is
the combined young and aged population per 100 economically active residents.

```{r}
#| label: age-groups
young_groups <- c("0_to_4", "5_to_9", "10_to_14", "15_to_19")

working_groups <- c(
  "20_to_24", "25_to_29", "30_to_34", "35_to_39", "40_to_44",
  "45_to_49", "50_to_54", "55_to_59", "60_to_64"
)

aged_groups <- c(
  "65_to_69", "70_to_74", "75_to_79", "80_to_84",
  "85_to_89", "90_and_over"
)
```

```{r}
#| label: prepare-population
population2024 <- population_raw |>
  group_by(PA, SZ, AG) |>
  summarise(Pop = sum(Pop, na.rm = TRUE), .groups = "drop") |>
  group_by(PA, SZ) |>
  summarise(
    YOUNG = sum(Pop[AG %in% young_groups]),
    ECONOMY_ACTIVE = sum(Pop[AG %in% working_groups]),
    AGED = sum(Pop[AG %in% aged_groups]),
    TOTAL = sum(Pop),
    .groups = "drop"
  ) |>
  mutate(
    DEPENDENCY = if_else(
      ECONOMY_ACTIVE > 0,
      100 * (YOUNG + AGED) / ECONOMY_ACTIVE,
      NA_real_
    ),
    PA_KEY = str_to_upper(str_squish(PA)),
    SZ_KEY = str_to_upper(str_squish(SZ))
  )

population2024 |>
  arrange(desc(TOTAL)) |>
  select(PA, SZ, YOUNG, ECONOMY_ACTIVE, AGED, TOTAL, DEPENDENCY) |>
  slice_head(n = 8)
```

The raw counts are rounded to the nearest ten by SingStat. I therefore avoid
presenting false precision, especially when comparing small subzones.

## 3. Joining attributes to geometry

The two datasets use different capitalization, so I create upper-case,
trimmed keys for planning area and subzone before joining. Using both keys is
safer than subzone name alone.

```{r}
#| label: join-data
mpsz_pop2024 <- mpsz2019 |>
  mutate(
    PA_KEY = str_to_upper(str_squish(PLN_AREA_N)),
    SZ_KEY = str_to_upper(str_squish(SUBZONE_N))
  ) |>
  left_join(population2024, by = c("PA_KEY", "SZ_KEY"))

join_check <- tibble(
  check = c("Boundary features", "Features with population", "Features without population"),
  value = c(
    nrow(mpsz_pop2024),
    sum(!is.na(mpsz_pop2024$TOTAL)),
    sum(is.na(mpsz_pop2024$TOTAL))
  )
)

join_check
```

Some planning subzones have no resident record. I keep them as `NA` rather
than changing them to zero: “no matched record” and “confirmed zero residents”
are not the same statement.

## 4. A first choropleth

`qtm()` gives me a fast diagnostic map. It is useful for checking the join and
spatial pattern before spending time on design.

```{r}
#| label: quick-map
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "Quick choropleth of total resident population in 2024."
tmap_mode("plot")
qtm(
  mpsz_pop2024,
  fill = "TOTAL",
  fill.scale = tm_scale_continuous(values = "brewer.blues"),
  fill.legend = tm_legend(title = "Residents")
)
```

The pattern is plausible, but a continuous colour ramp hides the decisions
behind class boundaries. I next compare two explicit classification methods.

## 5. Comparing classification methods

Quantiles place roughly the same number of subzones in each class. Equal
intervals preserve equal numeric widths, but a few high-population subzones can
leave most areas crowded into the lower classes.

```{r}
#| label: classification-comparison
#| fig-width: 10
#| fig-height: 5
#| fig-cap: "Dependency ratio classified with quantiles and equal intervals."
map_quantile <- tm_shape(mpsz_pop2024) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile", n = 5, values = "brewer.blues"
    ),
    fill.legend = tm_legend(title = "Dependants per 100")
  ) +
  tm_borders(col = "white", lwd = 0.4) +
  tm_title("Quantile")

map_equal <- tm_shape(mpsz_pop2024) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "equal", n = 5, values = "brewer.blues"
    ),
    fill.legend = tm_legend(title = "Dependants per 100")
  ) +
  tm_borders(col = "white", lwd = 0.4) +
  tm_title("Equal interval")

tmap_arrange(map_quantile, map_equal, ncol = 2)
```

For a national comparison I prefer quantiles because differences across the
full distribution remain visible. Equal intervals would be preferable if the
numeric distance between classes were the main message.

## 6. Refined map design

I keep one sequential palette because dependency ratio has a natural low-to-
high order. A thin white boundary separates adjacent subzones without
overpowering the fill, and the compass and scale bar make the map usable beyond
the immediate classroom context.

```{r}
#| label: refined-map
#| fig-width: 9
#| fig-height: 7
#| fig-cap: "Refined dependency-ratio choropleth for Singapore subzones."
dependency_breaks <- quantile(
  mpsz_pop2024$DEPENDENCY,
  probs = seq(0, 1, length.out = 6),
  na.rm = TRUE
) |>
  as.numeric() |>
  unique()

refined_map <- tm_shape(mpsz_pop2024) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      breaks = dependency_breaks,
      values = "brewer.blues"
    ),
    fill.legend = tm_legend(title = "Dependants per 100\nworking-age residents")
  ) +
  tm_borders(col = "white", lwd = 0.35) +
  tm_title("Resident dependency ratio, 2024") +
  tm_compass(type = "8star", position = tm_pos_in("left", "top")) +
  tm_scalebar(position = tm_pos_in("left", "bottom")) +
  tm_layout(frame = FALSE)

refined_map
```

## 7. Small multiples

The same colour scale should not be forced onto groups with very different
totals. I therefore create two maps with their own quantile breaks and arrange
them side by side.

```{r}
#| label: small-multiples
#| fig-width: 10
#| fig-height: 5
#| fig-cap: "Young and aged residents shown as coordinated small multiples."
map_young <- tm_shape(mpsz_pop2024) +
  tm_polygons(
    fill = "YOUNG",
    fill.scale = tm_scale_intervals(style = "quantile", n = 5, values = "brewer.greens"),
    fill.legend = tm_legend(title = "Age 0–19")
  ) +
  tm_borders(col = "white", lwd = 0.3) +
  tm_title("Young residents")

map_aged <- tm_shape(mpsz_pop2024) +
  tm_polygons(
    fill = "AGED",
    fill.scale = tm_scale_intervals(style = "quantile", n = 5, values = "brewer.purples"),
    fill.legend = tm_legend(title = "Age 65+")
  ) +
  tm_borders(col = "white", lwd = 0.3) +
  tm_title("Aged residents")

tmap_arrange(map_young, map_aged, ncol = 2)
```

Reading the maps together shows that “many older residents” is not simply the
opposite of “many young residents”. Several established housing areas have
large totals in both groups, so the dependency ratio remains valuable as a
relative measure.

## 8. Selection and a complementary chart

I define a transparent selection rule: the top 10% of populated subzones by
dependency ratio. The map locates them, while the bar chart makes the ranking
and values easier to compare.

```{r}
#| label: selection
dependency_cutoff <- quantile(
  mpsz_pop2024$DEPENDENCY,
  0.90,
  na.rm = TRUE
)

high_dependency <- mpsz_pop2024 |>
  filter(!is.na(DEPENDENCY), DEPENDENCY >= dependency_cutoff)

tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "#edf1f5", col = "white", lwd = 0.25) +
  tm_shape(high_dependency) +
  tm_polygons(fill = "#6f82aa", col = "#465d91", lwd = 0.7) +
  tm_title("Top 10% by dependency ratio")
```

```{r}
#| label: complementary-chart
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Highest dependency ratios among Singapore subzones."
high_dependency |>
  st_drop_geometry() |>
  arrange(DEPENDENCY) |>
  slice_tail(n = 12) |>
  mutate(SUBZONE_N = fct_reorder(SUBZONE_N, DEPENDENCY)) |>
  ggplot(aes(DEPENDENCY, SUBZONE_N)) +
  geom_col(fill = "#6f82aa") +
  labs(
    x = "Dependants per 100 working-age residents",
    y = NULL,
    title = "Subzones with the highest dependency ratios"
  ) +
  theme_minimal()
```

## 9. Interactive exploration

The static map communicates one main pattern; the interactive version supports
lookup. Hovering or clicking lets a reader inspect the subzone name, planning
area and demographic values without printing labels over the whole island.

```{r}
#| label: interactive-map
#| fig-height: 6
tmap_mode("view")

tm_shape(st_transform(mpsz_pop2024, 4326)) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(style = "quantile", n = 5, values = "brewer.blues"),
    fill.legend = tm_legend(title = "Dependency ratio"),
    popup.vars = c(
      "Subzone" = "SUBZONE_N",
      "Planning area" = "PLN_AREA_N",
      "Total residents" = "TOTAL",
      "Young residents" = "YOUNG",
      "Aged residents" = "AGED",
      "Dependency ratio" = "DEPENDENCY"
    )
  )

tmap_mode("plot")
```

## 10. What I learned

The join deserves as much attention as the map. Normalising both geographic
keys and retaining unmatched boundaries prevents silent data loss. I also found
that classification is an analytical decision: quantiles reveal relative
position, while equal intervals preserve numeric distance. Finally, pairing a
choropleth with a ranking chart makes the result easier to interpret than either
view alone.

## Reproducibility

The complete preparation workflow is available in
[`R/analysis.R`](R/analysis.R), and the source downloads are scripted in
[`R/download-data.R`](R/download-data.R).

```{r}
#| label: session-info
#| echo: false
sessionInfo()
```
