---
title: "Road Traffic Accidents in Greater Bangkok, 2022"
subtitle: "Technical report template · analysis pending"
description: "Uncompleted technical report structure for Take-home Exercise 1."
number-sections: true
code-tools: true
---

```{r}
#| label: thx-register-code
#| include: false
knitr::read_chunk("take home exercise1/R/prepare.R")
knitr::read_chunk("take home exercise1/R/analysis.R")
knitr::opts_chunk$set(fig.width = 9, fig.height = 6, dpi = 150)
```

::: {.callout-warning title="Draft — not ready for submission"}
The student has not yet supplied analytical code, maps, results, or interpretations.
The prompts below identify the work needed. Replace each prompt with your own
completed work before final publication.
:::

[Exercise home](index.qmd) · [Executive summary](executive-summary.qmd) ·
[Learning guide](learning-guide.qmd)

## Introduction and study context

::: {.th-prompt}
**Your contribution:** Explain the public-safety problem and the audience for
your report. Cite the evidence you use to establish context. Define the population
represented by this dataset without assuming that it captures every crash.
:::

## Research question and objectives

::: {.th-prompt}
**Your contribution:** State your main question, its spatial and temporal scope,
and what evidence would answer it. Distinguish the question about event intensity
from the question about interaction among events.
:::

## Data and study design

### Source inventory

| Input | Source and coverage | Acquisition record |
|---|---|---|
| Accident records | [Kaggle, Thailand Road Accident 2019–2022](https://www.kaggle.com/datasets/thaweewatboy/thailand-road-accident-2019-2022); nationwide source, 2022 study restriction | Downloaded 27 September 2026; original CSV and Parquet retained locally |
| Administrative boundaries | [geoBoundaries THA ADM1](https://www.geoboundaries.org/api/current/gbOpen/THA/ADM1/); 2017 vintage in source metadata | Boundary ID THA-ADM1-36821470; full-resolution GeoJSON retained locally |
| Road network, if used | Not acquired or selected | Document provider, extract date, historical suitability, license, and road coverage |

Checksums and original metadata are recorded in the [data notes](data/README.md).
Acquisition is complete; analytical validation of these inputs is still pending.

The RStudio workflow follows Exercise 2A. The report loads labelled sections of
[`R/prepare.R`](R/prepare.R) and [`R/analysis.R`](R/analysis.R) with
`knitr::read_chunk()`. Replace the corresponding script section with your own code,
then remove `eval: false` from its report chunk to execute it during rendering.
Do this in order so later sections can use objects created earlier.

```{r}
#| label: thx-packages
#| eval: false
```

```{r}
#| label: thx-import
#| eval: false
```

### Event definition

::: {.th-prompt}
**Your contribution:** Define one event. Explain how accident IDs, casualties,
repeated coordinates, and possible duplicate reports relate to that definition.
Record the evidence supporting your treatment of each.
:::

### Spatial and temporal windows

::: {.th-prompt}
**Your contribution:** Operationalise the proposed six-area extent with the
boundary dataset. Document the actual selected features and any boundary/name
mismatches. Explain the boundary vintage, incident-date interpretation, the 2022
filter, and any temporal grouping. Include your study-area map and caption.
:::

## Data preparation and quality assessment

::: {.th-prompt}
**Your contribution:** Insert your executed preparation code and explain each
major step. Document coordinate validation, CRS transformation, geometry checks,
temporal and spatial filters, and handling of missing or repeated records.
:::

| Check or transformation | Rule you applied | Removed or changed | Retained | Reason and consequence |
|---|---|---|---|---|
| Original import | Pending | — | — | Pending |
| Incident dates and 2022 filter | Pending | — | — | Pending |
| Duplicate reports | Pending | — | — | Pending |
| Missing or invalid coordinates | Pending | — | — | Pending |
| Study-area membership | Pending | — | — | Pending |
| Road matching, if applicable | Pending | — | — | Pending |

```{r}
#| label: thx-quality
#| eval: false
```

### Spatial preparation and study-area map

```{r}
#| label: thx-window
#| eval: false
```

```{r}
#| label: thx-study-map
#| eval: false
```

### Network preparation, if applicable

::: {.th-prompt}
**Your contribution:** Document road selection, topology, disconnected components,
grade-separated crossings, and how events were matched to roads. Report snapping
distances and exclusions. Explain how a network extract from a different year
might affect the study.
:::

```{r}
#| label: thx-network
#| eval: false
```

## Analytical assumptions and methodological decisions

| Decision | Your setting | Your justification | Assumption or sensitivity check |
|---|---|---|---|
| CRS and distance units | Pending | Pending | Pending |
| Event weights | Pending | Pending | Pending |
| Temporal aggregation | Pending | Pending | Pending |
| Kernel and bandwidth | Pending | Pending | Pending |
| Sampling spacing, if applicable | Pending | Pending | Pending |
| Distance range and step | Pending | Pending | Pending |
| Edge treatment | Pending | Pending | Pending |
| Reference/null model | Pending | Pending | Pending |
| Simulations, seed, and envelope type | Pending | Pending | Pending |

::: {.th-prompt}
Explain which assumptions are reasonable for your data and what departures would
mean. Cite methods you actually use. Parameters shown in a teaching example are
not a justification for your own settings.
:::

## First-order point-pattern analysis

::: {.th-prompt}
**Your contribution:** State the question, justify the method, show the executed
code and resulting maps or graphics, and explain their units. Discuss time where
it helps answer your question.

For each major visual, write your own interpretation of **no more than 150 words**:
what it shows, the supporting evidence, and its main limitation.
:::

```{r}
#| label: thx-first-order
#| eval: false
```

## Second-order point-pattern analysis

::: {.th-prompt}
**Your contribution:** State the question and null model before presenting the
result. Document assumptions, distance settings, edge treatment, simulation
settings, and the type of envelope or statistical test. Explain the result
relative to that model. Keep each major visual's interpretation within 150 words.
:::

```{r}
#| label: thx-second-order
#| eval: false
```

## Integrated findings and discussion

::: {.th-prompt}
**Your contribution:** Bring the first- and second-order findings together.
Explain agreements, differences, alternative explanations, and how reporting
coverage or exposure could affect conclusions. Insert your sensitivity results
and discuss whether the substantive conclusions change.
:::

```{r}
#| label: thx-sensitivity
#| eval: false
```

## Public-safety planning and management

::: {.th-prompt}
**Your contribution — maximum 500 words:** Link each proposed action to your
analytical evidence. Distinguish actions supported by the results from proposals
requiring traffic volumes, field assessment, or other information. Avoid causal
claims that your study cannot support.
:::

## Limitations

::: {.th-prompt}
**Your contribution:** Discuss the limitations actually relevant to your work,
including observation coverage, location accuracy, study-window choices,
network representation, parameter sensitivity, and inferential assumptions.
Explain what the analysis can and cannot establish.
:::

## Reproducibility

::: {.th-prompt}
**Your contribution:** Provide the script execution order, input checksums,
package versions, random seeds, and expected outputs. Include `sessionInfo()`
from your analytical run. The environment checker in this folder verifies setup
only; it does not establish that the analysis is reproducible.
:::

## References

These are starting references. Retain the ones used and add the sources supporting
your analytical choices and context.

- [Course assignment](https://isss626-ay2026-27aug.netlify.app/take-home_ex01)
- [Accident dataset](https://www.kaggle.com/datasets/thaweewatboy/thailand-road-accident-2019-2022)
- [geoBoundaries](https://www.geoboundaries.org/)
- [Bangkok and Vicinities definition](https://www.nso.go.th/public/e-book/Indicators-Environment/Environment-Indicators-2567/47/)
- [Chapter 7](https://r4gdsa.netlify.app/chap07)
- [spNetwork documentation](https://jeremygelb.github.io/spNetwork/reference/index.html)

## AI-use declaration

Generative AI helped prepare the website navigation, document and presentation
templates, learning notes, and source-acquisition records. It has not produced
analytical code, results, interpretations, or conclusions for this exercise.
The student must update this declaration to accurately describe the final work.
