---
title: "Hands-on Exercise 2B"
subtitle: "Second-order point patterns: G, F, K and L functions"
author: "Zhenhua Liu"
date: "2026-09-19"
body-classes: exercise-page
toc: true
toc-location: right
code-fold: false
code-tools: true
execute:
  freeze: false
  echo: true
  warning: true
  message: false
  error: false
---

```{r}
#| label: register-code
#| include: false
knitr::read_chunk("handson2A/R/prepare.R")
knitr::read_chunk("handson2B/R/analysis.R")
knitr::opts_chunk$set(fig.width = 9, fig.height = 5, dpi = 150)
```

## Overview

[Exercise 2A](../handson2A/index.qmd) mapped the intensity of childcare service
records. Here I ask a different question: how does the spacing of childcare
**locations** in Choa Chu Kang and Tampines compare with a uniform random
pattern inside the same planning boundaries?

Following [course Chapter 5](https://r4gdsa.netlify.app/chap05.html), I examine
G, F, K and L functions in that order, then compare observed curves with Monte
Carlo simulations. G and F describe nearest-neighbour and empty-space
distributions; K and L summarise cumulative interpoint relationships. These
descriptions can reveal departures from a reference model, but do not identify
the causal process that produced the locations.

## 1. Setup and reproducible preparation

```{r}
#| label: packages
```

I reuse the [ECDA childcare snapshot](https://data.gov.sg/datasets/d_5d668e3f544335f8028f546827b773b4/view)
and [MP2019 subzone boundary](https://data.gov.sg/datasets/d_8594ae9ff96d0c708bc2af633048edfb/view)
from 2A. The childcare layer describes **December 2021**, even though the portal
was updated in 2026. Reusing the input files makes the comparison temporally
consistent. It does not make the inventory current.

```{r}
#| label: import
```

```{r}
#| label: prepare-window
```

The shared preparation transforms both layers to EPSG:3414, checks geometry and
exclusions, and constructs polygonal observation windows. This page executes
that preparation itself; it does not depend on objects left in an earlier R session.

```{r}
#| label: planning-areas
```

::: {.callout-note title="Which events are being analysed?"}
Several childcare records share exactly the same coordinate. The principal
analysis here collapses these to distinct sites because a simple point-process
model assumes separate event locations. This is an explicit change of unit,
not a claim that the original records are erroneous. A final sensitivity analysis
restores every record. The record-based KDE in 2A remains unchanged.
:::

## 2. Study windows and the random reference model

Under the null model, the observed number of sites is fixed and each location is
independently uniform within its planning-area polygon. This is **conditional
complete spatial randomness (CSR)**. The alternative is departure from that
model in the selected distance range. Keeping the same boundary and count
avoids confounding location pattern with different totals or observation areas.

I preselect **0–500 m** and 129 equally spaced distances for both areas, before
examining the curves. This focuses the comparison on neighbourhood-scale
spacing and avoids searching many distance ranges for a favourable result.
Distances are straight-line metres, not walking routes. A 500 m separation
therefore has no automatic interpretation as a service catchment.

```{r}
#| label: second-order-setup
```

```{r}
#| label: area-map
#| fig-height: 6
#| fig-cap: "Distinct childcare sites and the actual observation windows. Each panel has its own map scale."
#| fig-alt: "Maps of distinct childcare sites in Choa Chu Kang and Tampines."
```

The analysis uses **`r npoints(patterns[['CHOA CHU KANG']])` sites** in Choa Chu
Kang and **`r npoints(patterns[['TAMPINES']])` sites** in Tampines. Both polygons
include land that is not equally plausible for childcare facilities. Rejection
of this homogeneous model could reflect land use or population gradients,
as well as genuine interaction between providers.

## 3. Simulation envelopes and edge correction

Points near an observation boundary may have unobserved neighbours outside it.
For G and F, I use Kaplan–Meier correction to handle censored distances. For K
and L, I use Ripley's isotropic edge correction. I apply the same estimator to
the observed pattern and every simulation. For G I also display the course's
border estimator as a sensitivity comparison.

The 199 simulated patterns in each area are reused across all four functions.
This makes differences between diagnostics easier to interpret because they
are not caused by unrelated simulation draws.

```{r}
#| label: envelope-method
```

::: {.callout-important title="An envelope is not automatically a global test"}
These plots show **95% pointwise** envelopes: with 199 simulations and
`nrank = 5`, the two-sided level at one preselected distance is
2 × 5 / 200 = 0.05. Scanning the whole curve for any crossing changes that
error rate. I therefore report a separate maximum-deviation Monte Carlo test
over the full 0–500 m interval and apply Holm adjustment across the eight
area/function comparisons.

The course's 99-simulation, rank-one pointwise envelope has level 0.02 at a
single distance, not 0.001. Our 199-simulation Monte Carlo p-value has resolution
0.005 and cannot support a claim of p &lt; 0.001. I use a stated 5% threshold
rather than repeating that mismatch.
:::

```{r}
#| label: plot-helper
#| code-fold: true
```

## 4. G: distance from one site to its nearest neighbour

G(r) is the probability that a typical site has another site within r metres.
Under homogeneous Poisson CSR its reference curve is
$G_0(r)=1-\exp(-\lambda\pi r^2)$, where $\lambda$ is sites per square metre.
An observed curve above the reference indicates more short nearest-neighbour
distances; below it indicates fewer. This direction describes spacing, not
proof that operators attract or repel one another.

```{r}
#| label: g-function
#| fig-cap: "G-function estimates and 95% pointwise simulation envelopes, using Kaplan–Meier correction."
#| fig-alt: "Nearest-neighbour distribution curves for Choa Chu Kang and Tampines against CSR envelopes."
```

The border-versus-Kaplan–Meier table is a reminder that edge treatment changes
the estimate, especially when many nearest-neighbour distances are censored.
The simulation plots use Kaplan–Meier consistently in both areas. A steep early
rise means many sites already have a close neighbour; levelling near one leaves
little additional information at larger distances.

## 5. F: distance from empty space to a site

F(r) starts at an arbitrary location in the study window and measures the
distance to the nearest site. A low F relative to CSR means more empty space
far from a site. Clusters can therefore produce a high G but a low F: sites
have close neighbours while the gaps between clusters remain large.

```{r}
#| label: f-function
#| fig-cap: "Empty-space functions with Kaplan–Meier correction and 95% pointwise envelopes."
#| fig-alt: "Empty-space distribution curves for the two planning areas."
```

Unlike G, F gives weight to the whole polygon rather than starting at the
observed childcare sites. Low-density industrial or open areas can therefore
strongly affect this diagnostic. The raster approximation used internally by
`Fest()` and the finite number of sites also limit precision. I do not interpret
F as the proportion of residents who can access childcare.

## 6. K: cumulative neighbours across distance

K(r) is the expected number of other events within r of a typical event,
normalised by intensity. Its CSR reference is $\pi r^2$ and its units here are
square metres. Values above the reference indicate an excess of neighbours
within that distance. K is cumulative: an excess established at small distances
can remain visible at larger distances.

```{r}
#| label: k-function
#| fig-cap: "Ripley-corrected K curves in m², with the CSR reference πr²."
#| fig-alt: "Cumulative K-function curves with simulation envelopes in both areas."
```

I plot K directly rather than subtracting r. Subtracting metres from square
metres is dimensionally inconsistent and does not centre the CSR expectation
at zero. The transformed L function below provides the appropriate centred
view in metre units.

## 7. L: a centred comparison with randomness

The transformation $L(r)=\sqrt{K(r)/\pi}$ has CSR reference r. Consequently,
**L(r) − r** has a zero reference: positive values indicate excess cumulative
neighbours and negative values indicate fewer. L repackages K for easier
reading; it is not an independent source of evidence.

```{r}
#| label: l-function
#| fig-cap: "Centred L curves; zero is the homogeneous CSR reference. Shading is pointwise, not a simultaneous band."
#| fig-alt: "Centred L-function curves for Choa Chu Kang and Tampines."
```

The height of L(r) − r is an effect-size description in metres, not a p-value.
I examine its direction alongside the explicit simulation results below,
rather than declaring significance from a visually prominent peak.

## 8. Global comparison and numerical interpretation

For each observed or simulated curve, I compute a single maximum absolute
deviation from the same CSR reference across the preselected distance grid.
The Monte Carlo p-value is $(1+b)/(1+199)$, where b is the number of simulated
deviations at least as large as the observed one. Counting ties as exceedances
is conservative. The reference is fixed by the area's count, geometry and
distance grid, so it is applied identically to all realisations.

```{r}
#| label: global-tests
```

After Holm adjustment, **`r sum(test_results$p_holm <= 0.05)` of the eight
comparisons** reject conditional CSR at 5%. The decision column is the formal
summary. Failure to reject means the chosen diagnostic and distance range do
not establish a departure; it does not prove that a pattern is random.
Because G, F, K and L are correlated, these are not eight independent votes.

To connect the curves to numbers, I also report their values at a preselected
**250 m** distance. This table describes effect direction; the pointwise bounds
are not substituted for the global p-values above.

```{r}
#| label: distance-summary
```

```{r}
#| label: computed-interpretation
#| echo: false
#| results: asis
for (nm in names(envelopes)) {
  e <- envelopes[[nm]]
  j <- which.min(abs(e$L$r - 250))
  delta <- e$L$obs[j] - e$L$r[j]
  cat(sprintf("**%s:** at 250 m, G is %.3f (CSR %.3f), F is %.3f (CSR %.3f), and L(r) − r is %.1f m. The centred L value indicates %s cumulative neighbours relative to the reference at this distance.\n\n",
    tools::toTitleCase(tolower(nm)), e$G$obs[j], e$G$theo[j], e$F$obs[j],
    e$F$theo[j], delta, if (delta > 0) "more" else "fewer"))
}
```

## 9. Sensitivity to co-located records

I repeat the L diagnostic with every original service record and a matching
fixed-count CSR simulation. The main site-based result and this record-based
result answer different questions. Their difference measures how much
co-location affects apparent clustering.

```{r}
#| label: duplicate-sensitivity
```

Co-located records add zero-distance pairs. The table shows the resulting
change in L(r) − r at 250 m instead of assuming that all apparent aggregation
comes from this issue. These additional unadjusted sensitivity p-values are
descriptive checks, separate from the predefined family of eight tests.

## 10. What I learned and limitations

Choosing the event unit came before choosing the function. Distinct childcare
sites and individual service records produce different nearest-neighbour
relationships, even though they appear almost identical on a national map.
G and F also showed why starting at a service location and starting at an
arbitrary piece of land are different perspectives.

The main limitation is the homogeneous reference. Childcare is unlikely to
locate uniformly across industrial, residential and open land. Rejecting CSR
does not separate this first-order heterogeneity from second-order interaction.
An inhomogeneous model with population or land-use covariates would be a
logical next step. Capacity, travel networks, historical boundary changes,
coordinate precision and cross-boundary access are also outside this analysis.

## Reproducibility and references

The executed sections are in [`R/analysis.R`](R/analysis.R); shared input
preparation is in [`../handson2A/R/prepare.R`](../handson2A/R/prepare.R).
Data sources and checksums are documented in
[`../handson2A/data/README.md`](../handson2A/data/README.md).
Run `source("handson2A/R/download-data.R")` once and then
`quarto render . --execute` from the repository root. Fixed seeds, 199
simulations, explicit corrections, and a declared distance grid make the
Monte Carlo comparisons reproducible within the recorded software environment.
The page fails if estimates are undefined on the requested grid.

References: [course Chapter 5](https://r4gdsa.netlify.app/chap05.html),
[spatstat](https://spatstat.org/), and the
[simulation-envelope documentation](https://search.r-project.org/CRAN/refmans/spatstat.explore/html/envelope.html).
The latter distinguishes pointwise and simultaneous testing and documents
`nrank`, `fix.n` and saved simulation functions.

```{r}
#| label: session-info
#| echo: false
sessionInfo()
```
