---
title: "Hands-on Exercise 2A"
subtitle: "First-order spatial point patterns and childcare intensity"
author: "Zhenhua Liu"
date: "2026-09-19"
body-classes: exercise-page
toc: true
toc-location: right
code-fold: false
code-tools: true
execute:
  freeze: true
  echo: true
  warning: true
  message: false
  error: false
---

```{r}
#| label: register-code
#| include: false
knitr::read_chunk("handson2A/R/prepare.R")
knitr::read_chunk("handson2A/R/analysis.R")
knitr::opts_chunk$set(fig.width = 9, fig.height = 6, dpi = 150)
```

## Overview

Where are childcare services concentrated, and how much does the answer depend
on the observation window and smoothing method? I use Singapore's childcare
locations to examine intensity at national and planning-area scales. A nearest-neighbour
test provides a benchmark against spatial randomness; kernel density estimation
(KDE) then shows where concentrations occur. These are complementary summaries,
not a measure of whether local demand is adequately served.

This exercise follows the workflow in [course Chapter 4](https://r4gdsa.netlify.app/chap04.html).
The next page, [Hands-on Exercise 2B](../handson2B/index.qmd), examines spatial
relationships over a range of distances.

## 1. Setup and source data

`sf` handles vector geometry, the `spatstat` component packages provide planar
point-pattern analysis, and `terra` carries the density raster into `tmap`.
`dplyr` supplies the attribute filtering used to define the study region.
Packages are installed before rendering, so the page never changes its environment
halfway through an analysis.

```{r}
#| label: packages
```

I use the official [ECDA Child Care Services GeoJSON](https://data.gov.sg/datasets/d_5d668e3f544335f8028f546827b773b4/view)
and [URA Master Plan 2019 Subzone Boundary (No Sea)](https://data.gov.sg/datasets/d_8594ae9ff96d0c708bc2af633048edfb/view).
The download routine is provided in [`R/download-data.R`](R/download-data.R).
`NAME` and `ADDRESSSTREETNAME` describe each service; `PLN_AREA_N` and `SUBZONE_N`
identify the administrative units.

::: {.callout-note title="The data date is not the download date"}
The ECDA portal describes this layer as **December 2021 data**, despite its
2026 portal update. These maps describe that historical service snapshot,
downloaded in September 2026. They should not be presented as a current inventory.
Both GeoJSON files already contain separate attributes, so no HTML-description
parsing is needed.
:::

```{r}
#| label: import
```

## 2. Geometry, CRS and the observation window

I transform both layers to **SVY21 / Singapore TM (EPSG:3414)** so coordinates
and distances are in metres. `st_transform()` changes coordinate values;
assigning a CRS label alone would not perform that conversion. I validate the
polygons and use the course's exclusions for the Southern Group, Western Islands
and North-Eastern Islands. This is a specified administrative study window,
not a claim that every remaining square metre is habitable land.

Dissolving subzone boundaries with `st_union()` produces one region for
`as.owin()`. An `owin` stores the observation boundary; a `ppp` stores events
inside that boundary. Using the points' bounding rectangle instead would include
unobserved sea and change both expected density and simulated patterns.

```{r}
#| label: prepare-window
```

The audit retains **`r npoints(childcare_ppp)` service records** at
**`r npoints(childcare_unique)` distinct coordinates**. There are
`r sum(duplicate_location)` records beyond the first at repeated coordinates and
`r sum(!inside)` points outside the window. Repeated coordinates are not enough
evidence to delete a service: several operators or records can share a building.
I retain records for provision intensity and also test distinct locations as a
sensitivity check. No random jitter is added to manufacture separation.

```{r}
#| label: national-map
#| fig-cap: "Historical childcare service records within the selected MP2019 study window."
#| fig-alt: "Map of Singapore subzones overlaid with childcare service points."
```

The overlay checks spatial alignment and reveals uneven concentrations. Large
low-density areas include land that is not primarily residential, so national
homogeneous randomness is a deliberately simple reference model.

## 3. Nearest-neighbour analysis

The Clark–Evans ratio compares observed mean nearest-neighbour distance with
the homogeneous Poisson expectation. Values below one indicate shorter
separations. I specify a one-sided clustering alternative for this national
question, with a 5% significance threshold.

I show both the asymptotic approximation and 999 Monte Carlo simulations.
The latter places the same number of points uniformly inside the actual
window and applies the same uncorrected statistic to observed and simulated
patterns. This accounts for window shape in the reference distribution, although
the uncorrected ratio itself still has edge bias. Nearest-neighbour statistics
involve interpoint distances; their inclusion here follows the course's sequence,
while KDE is the main first-order intensity method.

```{r}
#| label: nearest-neighbour
```

The record-level ratio is **`r round(unname(ce_mc$statistic), 3)`**, with simulated
p = **`r format(ce_mc$p.value, scientific = FALSE)`**. After collapsing co-located
records, R becomes **`r round(unname(ce_unique$statistic), 3)`** and p becomes
**`r format(ce_unique$p.value, scientific = FALSE)`**. The comparison separates
the contribution of shared coordinates from the broader geographic pattern.
With 999 simulations, the smallest attainable one-sided Monte Carlo p-value is
0.001; it is never zero.

For planning, a departure from uniform randomness means that service locations
are spatially uneven. It does not establish oversupply, undersupply, or a causal
attraction between providers. Residential distribution is an obvious omitted factor.

## 4. Bandwidth selection and density units

Bandwidth controls the scale at which a concentration is visible. I compute four
selectors rather than importing the numerical bandwidth printed in the lesson.
Scott's rule returns separate x and y smoothing scales, so its two values should
not be mistaken for two competing scalar solutions.

```{r}
#| label: bandwidths
```

The Diggle result is **`r round(as.numeric(bw_diggle) * 1000)` m**; likelihood
cross-validation gives **`r round(as.numeric(bw_ppl) * 1000)` m**. The table
documents how far alternative selectors move the scale of the question.

```{r}
#| label: kde-units
```

The two computations describe the same surface. Rescaling coordinates from
metres to kilometres multiplies density values by one million. I verify that
identity numerically, using matched 100 m cells. The raster integral is a useful
diagnostic against the input count; pixel discretisation and boundary correction
mean it need not reproduce the count exactly. Small negative values close to
machine precision are numerical convolution artefacts, not negative services.

```{r}
#| label: bandwidth-comparison
#| fig-cap: "Diggle and likelihood bandwidths, with a shared intensity scale in records per km². Map coordinates are kilometres."
#| fig-alt: "Two comparable childcare kernel density maps using different bandwidth selectors."
```

A narrower bandwidth preserves more local peaks; a wider bandwidth merges nearby
peaks into broader concentrations. I use the same colour limits so differences
are not created by separately stretched legends.

## 5. Kernel shape, fixed and adaptive smoothing

First I hold bandwidth constant and change the kernel. This isolates kernel
shape from bandwidth selection, with the same spatial resolution and intensity
scale throughout.

```{r}
#| label: kernel-comparison
#| fig-height: 8
#| fig-cap: "Gaussian, Epanechnikov, quartic and disc kernels at the same Diggle bandwidth; records per km²."
#| fig-alt: "Four childcare density surfaces comparing kernel shapes."
```

The compact-support kernels assign no contribution beyond their support, whereas
a Gaussian kernel decays gradually. Broad concentrations should be read across
the panels; an isolated bright pixel is not evidence of a statistically significant hotspot.

Next I compare the lesson's **600 m fixed bandwidth** with an adaptive kernel
estimate. The adaptive method allows local smoothing scales to vary with the
pilot intensity, which can retain detail where records are dense.

```{r}
#| label: adaptive-kde
#| fig-cap: "Fixed 600 m and adaptive kernel estimates on a shared records-per-km² scale."
#| fig-alt: "Side-by-side fixed and adaptive childcare density estimates."
```

Adaptive smoothing changes the estimator, so differences cannot be attributed
only to a single bandwidth value. Neither surface adjusts for population or
centre capacity. A high value describes many records per land area.

## 6. A geographically correct KDE map

::: {.callout-important title="Raster values and map coordinates use different units"}
EPSG:3414 requires metre coordinates. Assigning it to an image whose coordinates
were divided by 1,000 would put the raster in the wrong place. I convert the
**metre-coordinate image** to `SpatRaster`, multiply only its **density values**
by one million, and then attach EPSG:3414. The subzone overlay and scale bar are
therefore geographically consistent.
:::

```{r}
#| label: cartographic-kde
#| fig-height: 7
#| fig-cap: "Childcare service intensity, with correctly aligned MP2019 boundaries and a continuous sequential legend."
#| fig-alt: "Cartographic childcare kernel density map with boundaries, compass and scale bar."
```

I retain a continuous sequential palette because KDE is a continuous estimate.
Discretising it into quantiles would change the visual message to relative
ranks and hide the numeric gap between intensities.

## 7. Planning-area comparison

The four local windows are Punggol, Tampines, Choa Chu Kang and Jurong West.
Although the source boundary contains subzones, the comparisons below are at
**planning-area** level after dissolving the relevant subzones.

```{r}
#| label: planning-areas
```

```{r}
#| label: area-pattern-map
#| fig-height: 8
#| fig-cap: "Service records in the four planning-area windows; coordinates are kilometres."
#| fig-alt: "Childcare point patterns in Punggol, Tampines, Choa Chu Kang and Jurong West."
```

Choa Chu Kang has `r round(area_summary$records_per_km2[area_summary$planning_area == 'CHOA CHU KANG'], 1)`
records/km² compared with `r round(area_summary$records_per_km2[area_summary$planning_area == 'TAMPINES'], 1)`
in Tampines. This is an area-normalised comparison, not a ranking of service
adequacy. The actual distribution inside each polygon is still important.

```{r}
#| label: area-tests
```

Here the alternative is two-sided: either unusually short or unusually long
nearest-neighbour distances would be relevant. I explicitly select
`method = "MonteCarlo"`; supplying `nsim` alone does not change the default
asymptotic method. These two comparisons are exploratory.

```{r}
#| label: area-kde
#| fig-height: 8
#| fig-cap: "Planning-area KDE with area-specific Diggle bandwidths and common intensity limits; records per km²."
#| fig-alt: "Four planning-area childcare kernel density maps."
```

Each bandwidth is fitted within its own window, so fine detail is not equally
resolved in all panels. The common legend supports intensity comparison; a
common fixed bandwidth would be preferable for a strictly matched smoothing
comparison. Local boundary correction cannot represent cross-boundary access
to a nearby centre in another planning area.

## 8. Interactive lookup

The map supports centre-name and address lookup. External background tiles
require an internet connection; the analytical figures above are locally
rendered assets and remain available independently of the tile provider.

```{r}
#| label: interactive-points
```

## 9. What I learned

The observation window is part of the statistical model, not just a cartographic
frame. Counting service records and counting distinct locations are different
questions, especially when several records share a coordinate. I also found that
density units and coordinate units must be checked separately: a convincing
colour surface can still be geographically wrong if its extent is mislabelled.

The most useful next extension would relate service capacity to the residential
population of young children and realistic travel distances. This exercise alone
cannot identify unmet demand or explain why providers locate together.

## Reproducibility and references

All executed code is supplied in [`R/prepare.R`](R/prepare.R) and
[`R/analysis.R`](R/analysis.R). The page reads those labelled sections directly,
so the displayed code and executed code have one source. Run the download script
once, then `quarto render . --execute` from the repository root to rebuild the
whole site. The shared preparation also allows 2B to render independently.
Raw and derived data stay in ignored folders. [`data/README.md`](data/README.md)
records the sources, checksums and assumptions.

Method references: [course Chapter 4](https://r4gdsa.netlify.app/chap04.html),
[spatstat documentation](https://spatstat.org/),
[adaptive density reference](https://search.r-project.org/CRAN/refmans/spatstat.explore/html/adaptive.density.html),
and [sf coordinate transformations](https://r-spatial.github.io/sf/reference/st_transform.html).

```{r}
#| label: session-info
#| echo: false
sessionInfo()
```
