---
title: "Hands-on Exercise 1A"
subtitle: "Geospatial data wrangling with R"
author: "Zhenhua Liu"
date: "2026-09-11"
body-classes: exercise-page
toc: true
toc-location: right
code-fold: false
code-tools: true
execute:
  echo: true
  warning: false
  message: false
---

## Overview

This exercise is my first end-to-end spatial workflow in R. I import polygon,
line and point data, put every layer in a common projected coordinate reference
system, and then use spatial relationships to answer two practical questions:
how much land is covered by a cycling-path buffer in Tampines West, and how
pre-school provision varies between subzones.

::: {.callout-note title="Data note"}
The lesson was originally written around KML and shapefile downloads. The
current official portals provide GeoJSON, so I kept the analytical intention
but used the live formats available in September 2026. Raw files are obtained
by `R/download-data.R` and are not duplicated in the Git repository.
:::

## 1. Setup and data import

```{r}
#| label: setup
#| include: false
library(sf)
library(tidyverse)

data_dir <- "hands-on-exercise-1a/data/raw"
```

I first read one example of each geometry type. Looking at the geometry column
separately from the attribute table helped me remember that an `sf` object is
still a data frame; it simply carries a special list-column for shapes.

```{r}
#| label: import-spatial-data
mpsz <- st_read(file.path(data_dir, "mp14-subzone.geojson"), quiet = TRUE) |>
  st_make_valid()

preschool <- st_read(file.path(data_dir, "preschools.geojson"), quiet = TRUE)
cycling <- st_read(file.path(data_dir, "cycling-paths.geojson"), quiet = TRUE)

tibble(
  layer = c("Subzones", "Pre-schools", "Cycling paths"),
  features = c(nrow(mpsz), nrow(preschool), nrow(cycling)),
  geometry = c(
    as.character(unique(st_geometry_type(mpsz))[1]),
    as.character(unique(st_geometry_type(preschool))[1]),
    as.character(unique(st_geometry_type(cycling))[1])
  ),
  crs = c(st_crs(mpsz)$input, st_crs(preschool)$input, st_crs(cycling)$input)
)
```

```{r}
#| label: inspect-data
#| output: true
glimpse(st_drop_geometry(mpsz)[, 1:min(6, ncol(mpsz) - 1)])
head(st_geometry(mpsz), 2)
```

The default `plot()` method is useful as a quick diagnostic because it can show
both the feature geometry and individual attributes. I use one attribute here
to keep the result readable.

```{r}
#| label: quick-sf-plots
#| fig-width: 9
#| fig-height: 4.5
#| fig-cap: "Quick inspection of geometry, one attribute, and an overlay."
par(mfrow = c(1, 3), mar = c(0, 0, 2, 0))
plot(st_geometry(mpsz), col = "#dce7f3", border = "#ffffff", main = "Geometry")
plot(mpsz["SUBZONE_N"], key.pos = NULL, reset = FALSE, main = "Attribute")
plot(st_geometry(mpsz), col = "#eef3f8", border = "#aebcd0", main = "Overlay")
plot(st_geometry(preschool), add = TRUE, pch = 20, cex = 0.25, col = "#d45b5b")
par(mfrow = c(1, 1))
```

## 2. Coordinate reference systems

The imported files describe the same country but do not necessarily use the
same CRS. Latitude/longitude is suitable for storage and web display; distance
and area calculations need a projected CRS. I therefore transform every layer
to **SVY21 / Singapore TM (EPSG:3414)**.

```{r}
#| label: transform-crs
crs_before <- tibble(
  layer = c("Subzones", "Pre-schools", "Cycling paths"),
  epsg = c(st_crs(mpsz)$epsg, st_crs(preschool)$epsg, st_crs(cycling)$epsg)
)

mpsz <- st_transform(mpsz, 3414)
preschool <- st_transform(preschool, 3414)
cycling <- st_transform(cycling, 3414)

crs_before
```

::: {.callout-tip title="My check before analysis"}
`st_set_crs()` only changes the coordinate label. It does not move the
geometry. Because these layers already have valid coordinates, `st_transform()`
is the correct operation.
:::

## 3. Converting Airbnb records to spatial points

The Airbnb file is an ordinary CSV. Its longitude and latitude columns become
point geometry in WGS 84, after which I transform the points to EPSG:3414 for a
valid overlay with the planning boundaries.

```{r}
#| label: convert-airbnb
airbnb_raw <- read_csv(
  file.path(data_dir, "airbnb-listings.csv"),
  show_col_types = FALSE
)

airbnb <- airbnb_raw |>
  filter(!is.na(longitude), !is.na(latitude)) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(3414)

airbnb
```

```{r}
#| label: airbnb-overlay
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "Airbnb listings over Master Plan 2014 subzones."
ggplot() +
  geom_sf(data = mpsz, fill = "#eef3f8", colour = "white", linewidth = 0.15) +
  geom_sf(data = airbnb, colour = "#d45b5b", alpha = 0.45, size = 0.35) +
  coord_sf(datum = NA) +
  labs(title = "Airbnb listings in Singapore", subtitle = "Each red point is one listing") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

My first impression is that the listings are strongly clustered instead of
being evenly distributed. The central area and several mature neighbourhoods
stand out; that pattern would have been difficult to notice in the CSV alone.

## 4. Five-metre cycling-path buffer

I select Tampines West, intersect the cycling lines with its boundary, and
buffer the remaining lines by five metres. Summing every buffer polygon can
double-count overlaps, so I dissolve them before calculating the final area.

```{r}
#| label: cycling-buffer
tampines_west <- mpsz |>
  filter(SUBZONE_N == "TAMPINES WEST")

cycling_in_tampines <- st_intersection(cycling, tampines_west)
cycling_buffer <- st_buffer(cycling_in_tampines, dist = 5)
cycling_buffer_union <- st_union(cycling_buffer)

buffer_result <- tibble(
  measure = c("Sum before dissolving", "Dissolved footprint"),
  area_m2 = c(
    sum(as.numeric(st_area(cycling_buffer))),
    as.numeric(st_area(cycling_buffer_union))
  )
) |>
  mutate(area_km2 = area_m2 / 1e6)

buffer_result
```

```{r}
#| label: cycling-buffer-map
#| fig-width: 8
#| fig-height: 5.5
#| fig-cap: "Dissolved five-metre buffer around cycling paths in Tampines West."
ggplot() +
  geom_sf(data = tampines_west, fill = "#eef3f8", colour = "#465d91") +
  geom_sf(data = cycling_buffer_union, fill = "#73c6b6", colour = NA, alpha = 0.85) +
  geom_sf(data = cycling_in_tampines, colour = "#24435d", linewidth = 0.35) +
  coord_sf(datum = NA) +
  labs(title = "Cycling-path buffer", subtitle = "Tampines West, 5 metres") +
  theme_void()
```

The dissolved result is the more defensible estimate of land covered. The gap
between the two numbers is a useful reminder that geometry overlaps are not
just a display issue—they change summary statistics.

## 5. Pre-school count and density

`st_intersects()` returns the pre-schools falling inside each subzone. I keep
both the count and the density because they answer different questions: count
describes total provision, while density adjusts for the physical size of the
subzone.

```{r}
#| label: preschool-density
mpsz_summary <- mpsz |>
  mutate(
    preschool_count = lengths(st_intersects(mpsz, preschool)),
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    preschool_density = preschool_count / area_km2
  )

highest_count <- mpsz_summary |>
  st_drop_geometry() |>
  slice_max(preschool_count, n = 5, with_ties = FALSE) |>
  select(SUBZONE_N, PLN_AREA_N, preschool_count, area_km2, preschool_density)

highest_count
```

```{r}
#| label: preschool-map
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "Pre-school locations and subzone-level counts."
ggplot(mpsz_summary) +
  geom_sf(aes(fill = preschool_count), colour = "white", linewidth = 0.12) +
  geom_sf(data = preschool, colour = "#34495e", alpha = 0.45, size = 0.22) +
  scale_fill_viridis_c(option = "C", direction = -1) +
  coord_sf(datum = NA) +
  labs(title = "Pre-schools by subzone", fill = "Count") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

```{r}
#| label: preschool-diagnostics
#| fig-width: 9
#| fig-height: 4
#| fig-cap: "Distribution of density and the relationship between count and density."
density_data <- mpsz_summary |>
  st_drop_geometry() |>
  filter(is.finite(preschool_density))

p_hist <- ggplot(density_data, aes(preschool_density)) +
  geom_histogram(bins = 25, fill = "#6f82aa", colour = "white") +
  labs(x = "Pre-schools per km²", y = "Subzones", title = "Density distribution") +
  theme_minimal()

p_scatter <- ggplot(
  density_data,
  aes(preschool_count, preschool_density)
) +
  geom_point(colour = "#465d91", alpha = 0.65) +
  geom_smooth(method = "lm", se = FALSE, colour = "#d45b5b", linewidth = 0.7) +
  labs(x = "Pre-school count", y = "Pre-schools per km²", title = "Count versus density") +
  theme_minimal()

gridExtra::grid.arrange(p_hist, p_scatter, ncol = 2)
```

The scatterplot confirms that a high count does not automatically imply the
highest density. Small subzones can rank highly after area is taken into
account. If I were extending this analysis, I would next relate provision to
the number of young children rather than land area alone.

## 6. What I learned

Three checks made the workflow more reliable: inspect geometry and attributes
before analysis, transform rather than relabel a valid CRS, and dissolve
overlapping buffers before measuring area. The exercise also showed me why a
map and a statistical plot should be read together—the map reveals geography,
while the histogram and scatterplot reveal the shape of the values.

## Reproducibility

The complete analysis is also available as
[`R/analysis.R`](R/analysis.R), and the data download steps are in
[`R/download-data.R`](R/download-data.R). Run the download script once, then
render this page from the project root.

```{r}
#| label: session-info
#| echo: false
sessionInfo()
```
